---
title: "04a - Multinomial (Conditional) Logit"
author: "Team"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(mlogit)
library(here)
source(here("R", "log_loss.R"))   # log_loss(actual, pred): both n x 4 matrices
```

Multinomial (conditional) logit following the class structure, with a
respondent-level split and log-loss scoring. Generates a submission and logs it.

# Load & derive Choice

```{r load}
train <- read.csv(here("data", "train.csv"))
test  <- read.csv(here("data", "test.csv"))

train$Choice <- max.col(train[, c("Ch1","Ch2","Ch3","Ch4")])   # 1..4
test$Choice  <- 1L                                             # placeholder for dfidx
```

# Respondent-level split

```{r split}
set.seed(7402)
cases     <- unique(train$Case)
val_cases <- sample(cases, size = round(0.20 * length(cases)))

train_fit <- subset(train, !(Case %in% val_cases))
train_val <- subset(train,   Case %in% val_cases)
```

# Reshape (dfidx, wide -> long)

```{r dfidx}
S_fit <- dfidx(train_fit, shape = "wide", choice = "Choice",
               varying = c(4:83), sep = "", idx = c("No", "Case"))
S_val <- dfidx(train_val, shape = "wide", choice = "Choice",
               varying = c(4:83), sep = "", idx = c("No", "Case"))
```

# Fit MNL

We include alternative-specific intercepts (they capture the opt-out's baseline
pull and improved our log-loss vs the no-intercept `- 1` spec).

```{r fit}
attr_rhs <- "CC+GN+NS+BU+FA+LD+BZ+FC+FP+RP+PP+KA+SC+TS+NV+MA+LB+AF+HU+Price"
f_mnl <- as.formula(paste("Choice ~", attr_rhs))   # intercepts kept

M <- mlogit(f_mnl, data = S_fit)
summary(M)
```

# Validation log loss

```{r logloss}
val_prob  <- predict(M, newdata = S_val)
val_truth <- train_val$Choice
truth_mat <- matrix(0, length(val_truth), 4)
truth_mat[cbind(seq_along(val_truth), val_truth)] <- 1

ll_mnl <- log_loss(truth_mat, val_prob)
cat("MNL validation log loss:", round(ll_mnl, 5), " (benchmark 1.38629)\n")
```

# Generate submission

```{r submission}
S_all  <- dfidx(train, shape = "wide", choice = "Choice",
                varying = c(4:83), sep = "", idx = c("No", "Case"))
S_test <- dfidx(test,  shape = "wide", choice = "Choice",
                varying = c(4:83), sep = "", idx = c("No", "Case"))

M_full    <- mlogit(f_mnl, data = S_all)
test_prob <- predict(M_full, newdata = S_test)

sample_sub <- read.csv(here("data", "sample_submission.csv"))
submission <- data.frame(No = test$No,
                         Ch1 = test_prob[,1], Ch2 = test_prob[,2],
                         Ch3 = test_prob[,3], Ch4 = test_prob[,4])
stopifnot(all(names(submission) == names(sample_sub)),
          all(submission$No == sample_sub$No), !anyNA(submission))

if (!dir.exists(here("submissions"))) dir.create(here("submissions"))
write.csv(submission, here("submissions", "submission_mnl.csv"), row.names = FALSE)
cat("Wrote submissions/submission_mnl.csv\n")
```

# Log the submission

```{r log}
path <- here("submissions_log.csv")
log  <- read.csv(path, stringsAsFactors = FALSE)
new  <- data.frame(
  date              = as.character(Sys.Date()),
  model_name        = "mnl_conditional_logit",
  description       = "Multinomial (conditional) logit with ASCs; all 19 attributes + Price, generic linear coefficients",
  source_citation   = "Class material (mlogit / dfidx)",
  cv_logloss        = round(ll_mnl, 6),
  public_lb_logloss = NA, gap = NA,
  notes             = "Class-structure MNL, respondent-level split, log-loss scored.",
  stringsAsFactors  = FALSE
)
write.csv(rbind(log, new), path, row.names = FALSE)
cat("Logged mnl_conditional_logit\n")
```
