---
title: "04b - Mixed Logit (random coefficients)"
author: "Team"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(mlogit)
library(here)
source(here("R", "log_loss.R"))
```

Mixed logit following the class notebook: coefficients treated as random
(normal) across respondents, exploiting the panel (19 tasks/person). Same
respondent-level split and log-loss scoring. Generates a submission and logs it.

> Note: mixed logit is slower than MNL. Start with a few random coefficients
> and a modest number of draws (`R`); raise `R` for the final fit.

# Load & derive Choice

```{r load}
train <- read.csv(here("data", "train.csv"))
test  <- read.csv(here("data", "test.csv"))

train$Choice <- max.col(train[, c("Ch1","Ch2","Ch3","Ch4")])
test$Choice  <- 1L
```

# Respondent-level split

```{r split}
set.seed(7402)
cases     <- unique(train$Case)
val_cases <- sample(cases, size = round(0.20 * length(cases)))

train_fit <- subset(train, !(Case %in% val_cases))
train_val <- subset(train,   Case %in% val_cases)
```

# Reshape — NOTE the list() idx form for panel/mixed logit

For the mixed logit the class reads the index as a **list**:
`idx = list(c("No","Case"))`. This is what lets `panel = TRUE` work.

```{r dfidx}
S_fit <- dfidx(train_fit, shape = "wide", choice = "Choice",
               varying = c(4:83), sep = "", idx = list(c("No", "Case")))
S_val <- dfidx(train_val, shape = "wide", choice = "Choice",
               varying = c(4:83), sep = "", idx = list(c("No", "Case")))
```

# Fit mixed logit

`rpar` marks coefficients as random normal (`"n"`). The class version makes ALL
coefficients random; that's the most flexible but slowest. Start here with all
random (matching class), reduce if it's too slow.

```{r fit}
attr_vars <- c("CC","GN","NS","BU","FA","LD","BZ","FC","FP","RP",
               "PP","KA","SC","TS","NV","MA","LB","AF","HU","Price")
rpar_all <- setNames(rep("n", length(attr_vars)), attr_vars)

f_mix <- as.formula(paste("Choice ~", paste(attr_vars, collapse = "+"), "- 1"))

M1 <- mlogit(f_mix, data = S_fit,
             rpar = rpar_all, panel = TRUE, R = 50, halton = NA)
summary(M1)
```

# Validation log loss

```{r logloss}
val_prob  <- predict(M1, newdata = S_val)
val_truth <- train_val$Choice
truth_mat <- matrix(0, length(val_truth), 4)
truth_mat[cbind(seq_along(val_truth), val_truth)] <- 1

ll_mix <- log_loss(truth_mat, val_prob)
cat("Mixed logit validation log loss:", round(ll_mix, 5),
    " (benchmark 1.38629)\n")
```

# Model comparison (class-style: log-likelihood & AIC)

```{r compare}
cat("Mixed logit log-likelihood:", round(as.numeric(logLik(M1)), 2), "\n")
cat("Mixed logit AIC:", round(AIC(M1), 2), "\n")
# A significant sd.* on a coefficient => preference heterogeneity for that
# feature across respondents (a reportable insight).
```

# Generate submission

```{r submission}
S_all  <- dfidx(train, shape = "wide", choice = "Choice",
                varying = c(4:83), sep = "", idx = list(c("No", "Case")))
S_test <- dfidx(test,  shape = "wide", choice = "Choice",
                varying = c(4:83), sep = "", idx = list(c("No", "Case")))

M1_full   <- mlogit(f_mix, data = S_all,
                    rpar = rpar_all, panel = TRUE, R = 50, halton = NA)
test_prob <- predict(M1_full, newdata = S_test)

sample_sub <- read.csv(here("data", "sample_submission.csv"))
submission <- data.frame(No = test$No,
                         Ch1 = test_prob[,1], Ch2 = test_prob[,2],
                         Ch3 = test_prob[,3], Ch4 = test_prob[,4])
stopifnot(all(names(submission) == names(sample_sub)),
          all(submission$No == sample_sub$No), !anyNA(submission))

if (!dir.exists(here("submissions"))) dir.create(here("submissions"))
write.csv(submission, here("submissions", "submission_mixed.csv"), row.names = FALSE)
cat("Wrote submissions/submission_mixed.csv\n")
```

# Log the submission

```{r log}
path <- here("submissions_log.csv")
log  <- read.csv(path, stringsAsFactors = FALSE)
new  <- data.frame(
  date              = as.character(Sys.Date()),
  model_name        = "mixed_logit_all_random",
  description       = "Mixed logit: all 19 attributes + Price as random-normal coefficients, panel over respondents, R=50 draws",
  source_citation   = "Class material (mlogit rpar/panel mixed logit)",
  cv_logloss        = round(ll_mix, 6),
  public_lb_logloss = NA, gap = NA,
  notes             = "Class-structure mixed logit. Raise R for final fit. Significant sd.* => preference heterogeneity.",
  stringsAsFactors  = FALSE
)
write.csv(rbind(log, new), path, row.names = FALSE)
cat("Logged mixed_logit_all_random\n")
```
