---
title: "05 - Improvements: what each change does to log-loss"
author: "Team"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(mlogit)
library(here)
source(here("R", "log_loss.R"))
```

Every model is scored on the SAME respondent-level split (seed 7402), so the
log-losses are comparable. Each section states the change and its effect; a
summary table and an experiments log are written at the end.

Note: attribute *factoring* is not included here — it fits the long-format
workflow, not the wide/`dfidx` reshape used in class. From our earlier
long-format run it scored ~1.2263; cite that separately if needed.

# Setup (shared split, helper)

```{r prep}
train <- read.csv(here("data", "train.csv"))
test  <- read.csv(here("data", "test.csv"))
train$Choice <- max.col(train[, c("Ch1","Ch2","Ch3","Ch4")])
test$Choice  <- 1L

set.seed(7402)
cases     <- unique(train$Case)
val_cases <- sample(cases, size = round(0.20 * length(cases)))
train_fit <- subset(train, !(Case %in% val_cases))
train_val <- subset(train,   Case %in% val_cases)

val_truth <- train_val$Choice
truth_mat <- matrix(0, length(val_truth), 4)
truth_mat[cbind(seq_along(val_truth), val_truth)] <- 1

attrs <- c("CC","GN","NS","BU","FA","LD","BZ","FC","FP","RP",
           "PP","KA","SC","TS","NV","MA","LB","AF","HU")

mk <- function(df, panel = FALSE) {
  idx <- if (panel) list(c("No","Case")) else c("No","Case")
  dfidx(df, shape = "wide", choice = "Choice",
        varying = c(4:83), sep = "", idx = idx)
}

results  <- list()
val_pred <- list()
```

# 0. MNL baseline (linear attrs + ASCs)

**Change:** none — reference model.

```{r m0}
f0 <- as.formula(paste("Choice ~", paste(c(attrs,"Price"), collapse="+")))
M0 <- mlogit(f0, data = mk(train_fit))
p0 <- predict(M0, newdata = mk(train_val))
val_pred[["mnl"]] <- p0
results[["0_MNL_baseline"]] <- list(ll = log_loss(truth_mat, p0),
                                    change = "reference: linear attrs + ASCs")
cat("0 MNL baseline:", round(results[["0_MNL_baseline"]]$ll,5), "\n")
```

# 1. Price non-linearity (add Price^2)

**Change:** allow a curved price effect (EDA showed choice rate falling steeply
then flattening). Adds one term.

```{r m1}
f1 <- as.formula(paste("Choice ~", paste(c(attrs,"Price"), collapse="+"),
                       "+ I(Price^2)"))
M1 <- mlogit(f1, data = mk(train_fit))
p1 <- predict(M1, newdata = mk(train_val))
val_pred[["price2"]] <- p1
results[["1_price_squared"]] <- list(ll = log_loss(truth_mat, p1),
                                     change = "added I(Price^2) curvature")
cat("1 price^2:", round(results[["1_price_squared"]]$ll,5), "\n")
```

# 2. Income x Price interaction

**Change:** price sensitivity depends on income (richer respondents less
price-sensitive). One interaction term with a clear real-world story.

```{r m2}
tf <- train_fit; tv <- train_val
for (j in 1:4) {
  tf[[paste0("PxI", j)]] <- tf[[paste0("Price", j)]] * tf$incomea
  tv[[paste0("PxI", j)]] <- tv[[paste0("Price", j)]] * tv$incomea
}
S2f <- dfidx(tf, shape="wide", choice="Choice", sep="",
             varying = c(4:83, which(names(tf) %in% paste0("PxI",1:4))),
             idx = c("No","Case"))
S2v <- dfidx(tv, shape="wide", choice="Choice", sep="",
             varying = c(4:83, which(names(tv) %in% paste0("PxI",1:4))),
             idx = c("No","Case"))
f2  <- as.formula(paste("Choice ~", paste(c(attrs,"Price","PxI"), collapse="+")))
M2  <- mlogit(f2, data = S2f)
p2  <- predict(M2, newdata = S2v)
val_pred[["income_price"]] <- p2
results[["2_income_x_price"]] <- list(ll = log_loss(truth_mat, p2),
                                      change = "Price interacts with income")
cat("2 income x price:", round(results[["2_income_x_price"]]$ll,5), "\n")
```

# 3. Nested logit (real bundles vs opt-out)

**Change:** treat the 3 real bundles as a nest separate from the opt-out —
"buy something vs buy nothing" as a distinct decision. More realistic
substitution than plain MNL.

```{r m3}
M3 <- tryCatch(
  mlogit(f0, data = mk(train_fit),
         nests = list(buy = c("1","2","3"), noBuy = c("4")),
         un.nest.el = TRUE),
  error = function(e) { message("Nested failed: ", conditionMessage(e)); NULL })
if (!is.null(M3)) {
  p3 <- predict(M3, newdata = mk(train_val))
  val_pred[["nested"]] <- p3
  results[["3_nested_logit"]] <- list(ll = log_loss(truth_mat, p3),
                                      change = "nest: 3 real bundles vs opt-out")
} else results[["3_nested_logit"]] <- list(ll = NA, change = "nested (failed)")
cat("3 nested:", round(results[["3_nested_logit"]]$ll,5), "\n")
```

# 4. Mixed logit (random Price)

**Change:** price coefficient varies across respondents (panel). Captures
heterogeneity fixed models can't.

```{r m4}
M4 <- tryCatch(
  mlogit(as.formula(paste("Choice ~", paste(c(attrs,"Price"),collapse="+"), "- 1")),
         data = mk(train_fit, panel = TRUE),
         rpar = c(Price = "n"), panel = TRUE, R = 50, halton = NA),
  error = function(e) { message("Mixed failed: ", conditionMessage(e)); NULL })
if (!is.null(M4)) {
  p4 <- predict(M4, newdata = mk(train_val, panel = TRUE))
  val_pred[["mixed"]] <- p4
  results[["4_mixed_logit"]] <- list(ll = log_loss(truth_mat, p4),
                                     change = "random Price coefficient (panel)")
} else results[["4_mixed_logit"]] <- list(ll = NA, change = "mixed (failed)")
cat("4 mixed:", round(results[["4_mixed_logit"]]$ll,5), "\n")
```

# 5. Ensemble (average the models)

**Change:** average predicted probabilities across models. Because models make
different errors, the average is often better-calibrated than any single one.

```{r m5}
pick <- intersect(c("mnl","price2","income_price","mixed"), names(val_pred))
ens  <- Reduce(`+`, val_pred[pick]) / length(pick)
ens  <- ens / rowSums(ens)
results[["5_ensemble"]] <- list(ll = log_loss(truth_mat, ens),
                                change = paste("avg of:", paste(pick, collapse=", ")))
cat("5 ensemble:", round(results[["5_ensemble"]]$ll,5), "\n")
```

# Summary: change vs log-loss

```{r summary}
summary_tbl <- data.frame(
  model      = names(results),
  change     = sapply(results, function(x) x$change),
  cv_logloss = round(sapply(results, function(x) x$ll), 5),
  row.names  = NULL
)
summary_tbl <- summary_tbl[order(summary_tbl$cv_logloss), ]
print(summary_tbl)
cat("\nbest:", round(min(summary_tbl$cv_logloss, na.rm=TRUE), 5), "\n")
```

# Log ALL experiments (monitoring record, separate from submissions_log)

```{r log-experiments}
exp_path <- here("model_experiments_log.csv")
run_tbl  <- summary_tbl
run_tbl$date <- as.character(Sys.Date())
run_tbl <- run_tbl[, c("date","model","change","cv_logloss")]
if (file.exists(exp_path)) {
  prev <- read.csv(exp_path, stringsAsFactors = FALSE)
  write.csv(rbind(prev, run_tbl), exp_path, row.names = FALSE)
} else write.csv(run_tbl, exp_path, row.names = FALSE)
cat("Appended", nrow(run_tbl), "experiments to model_experiments_log.csv\n")
```

# Generate + log a submission for the winning model

Pick the winner from the summary table. Default below = the ensemble (usually
the best). Only upload if it beats your current best on the public board.

```{r submit-winner}
# Refit each component on ALL training data, average test predictions.
train$Choice <- max.col(train[, c("Ch1","Ch2","Ch3","Ch4")])

# MNL on full data
Mf_mnl  <- mlogit(f0, data = mk(train))
tp_mnl  <- predict(Mf_mnl, newdata = mk(test))

# Price^2 on full data
Mf_p2   <- mlogit(f1, data = mk(train))
tp_p2   <- predict(Mf_p2, newdata = mk(test))

# Mixed on full data
Mf_mix  <- tryCatch(
  mlogit(as.formula(paste("Choice ~", paste(c(attrs,"Price"),collapse="+"), "- 1")),
         data = mk(train, panel = TRUE),
         rpar = c(Price = "n"), panel = TRUE, R = 50, halton = NA),
  error = function(e) NULL)
tp_mix  <- if (!is.null(Mf_mix)) predict(Mf_mix, newdata = mk(test, panel = TRUE)) else NULL

# ensemble available components
comp <- list(tp_mnl, tp_p2, tp_mix)
comp <- comp[!sapply(comp, is.null)]
tp   <- Reduce(`+`, comp) / length(comp)
tp   <- tp / rowSums(tp)

sample_sub <- read.csv(here("data", "sample_submission.csv"))
submission <- data.frame(No = test$No, Ch1=tp[,1], Ch2=tp[,2], Ch3=tp[,3], Ch4=tp[,4])
stopifnot(all(names(submission)==names(sample_sub)),
          all(submission$No==sample_sub$No), !anyNA(submission))
if (!dir.exists(here("submissions"))) dir.create(here("submissions"))
write.csv(submission, here("submissions","submission_ensemble.csv"), row.names = FALSE)

path <- here("submissions_log.csv"); log <- read.csv(path, stringsAsFactors = FALSE)
new  <- data.frame(date=as.character(Sys.Date()), model_name="ensemble_mnl_price2_mixed",
  description="Average of MNL, MNL+Price^2, and mixed-logit(random Price) probabilities",
  source_citation="Class material (mlogit); simple probability averaging",
  cv_logloss=round(results[["5_ensemble"]]$ll,6),
  public_lb_logloss=NA, gap=NA,
  notes="Ensemble of choice models. See 05 summary for component scores.",
  stringsAsFactors=FALSE)
write.csv(rbind(log, new), path, row.names = FALSE)
cat("Wrote + logged submission_ensemble.csv\n")
```
