---
title: "MNL improvements - stability comparison (mean, sd, min, max)"
output: html_document
date: "2026-07-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Each MNL variant is run over 20 respondent-level splits (matching the test
set's unseen respondents). For each we report mean / sd / min / max log-loss, so
we keep only changes that lower the mean WITHOUT inflating the spread.

Fix vs previous version: engineered columns must have NO separator before the
alternative number (dfidx `sep=""` expects `P21`, not `P2_1`).

# Setup

```{r}
library(mlogit)
train <- read.csv("/Users/imeldalee/Desktop/TAE KAGGLE/kagglecomp/data/train.csv")
train$Choice <- max.col(train[, c("Ch1","Ch2","Ch3","Ch4")])

log_loss <- function(actual, pred, eps = 1e-15) {
  actual <- as.matrix(actual); pred <- as.matrix(pred)
  stopifnot(all(dim(actual) == dim(pred)))
  pred <- pred / rowSums(pred)
  pred <- pmin(pmax(pred, eps), 1 - eps)
  -mean(rowSums(actual * log(pred)))
}
one_hot <- function(v) { m <- matrix(0, length(v), 4); m[cbind(seq_along(v), v)] <- 1; m }

attrs    <- c("CC","GN","NS","BU","FA","LD","BZ","FC","FP","RP",
              "PP","KA","SC","TS","NV","MA","LB","AF","HU")
attr_rhs <- paste(c(attrs, "Price"), collapse = "+")
```

# One function: fit a chosen variant on one split, return log-loss

```{r}
# variant: "baseline", "logprice", "price2", "income_price", "factors"
run_once <- function(seed, variant) {
  set.seed(seed)
  vc  <- sample(unique(train$Case), round(0.20 * length(unique(train$Case))))
  fit <- subset(train, !(Case %in% vc))
  val <- subset(train,   Case %in% vc)
  truth <- one_hot(val$Choice)

  # ---- engineered columns (NO underscore before the alt number) ----
  add_cols <- function(df) {
    if (variant == "income_price")
      for (j in 1:4) df[[paste0("PxI", j)]] <- df[[paste0("Price", j)]] * df$incomea
    if (variant == "logprice")
      for (j in 1:4) df[[paste0("lP", j)]]  <- log1p(df[[paste0("Price", j)]])
    if (variant == "price2")
      for (j in 1:4) df[[paste0("P2", j)]]  <- df[[paste0("Price", j)]]^2
    df
  }
  fit <- add_cols(fit); val <- add_cols(val)

  # ---- factors variant: collapse rare levels in the wide columns first ----
  if (variant == "factors") {
    ftr <- c("NS","BU","FP","SC","MA","LB")
    for (a in ftr) for (j in 1:4) {
      col <- paste0(a, j)
      tab <- table(fit[[col]]); keep <- as.numeric(names(tab)[tab >= 30])
      modal <- as.numeric(names(which.max(tab)))
      fit[[col]][!(fit[[col]] %in% keep)] <- modal
      val[[col]][!(val[[col]] %in% keep)] <- modal
    }
  }

  # ---- which wide columns are the alternative-varying ones ----
  extra <- switch(variant,
                  income_price = paste0("PxI", 1:4),
                  logprice     = paste0("lP", 1:4),
                  price2       = paste0("P2", 1:4),
                  character(0))
  vary_names <- c(names(fit)[4:83], extra)

  Sf <- dfidx(fit, shape="wide", choice="Choice", sep="",
              varying = which(names(fit) %in% vary_names), idx = c("No","Case"))
  Sv <- dfidx(val, shape="wide", choice="Choice", sep="",
              varying = which(names(val) %in% vary_names), idx = c("No","Case"))

  # ---- factor the selected attributes AFTER reshape (factors variant) ----
  if (variant == "factors") {
    ftr <- c("NS","BU","FP","SC","MA","LB")
    for (a in ftr) {
      Sf[[a]] <- factor(Sf[[a]])
      Sv[[a]] <- factor(Sv[[a]], levels = levels(Sf[[a]]))
    }
  }

  # ---- formula ----
  rhs <- attr_rhs
  if (variant == "income_price") rhs <- paste(rhs, "+ PxI")
  if (variant == "logprice")     rhs <- paste(rhs, "+ lP")
  if (variant == "price2")       rhs <- paste(rhs, "+ P2")

  M <- mlogit(as.formula(paste("Choice ~", rhs, "- 1")), data = Sf)
  p <- predict(M, newdata = Sv)
  log_loss(truth, p)
}
```

# Run each variant over 20 splits and collect the 5 stats

```{r}
variants <- c("baseline","logprice","price2","income_price","factors")
seeds    <- 1:20

summarise_variant <- function(v) {
  lls <- sapply(seeds, function(s) tryCatch(run_once(s, v), error = function(e) NA))
  lls <- lls[!is.na(lls)]
  if (length(lls) == 0)
    return(data.frame(variant=v, n=0, mean_ll=NA, sd_ll=NA, min_ll=NA, max_ll=NA))
  data.frame(variant = v,
             n       = length(lls),
             mean_ll = round(mean(lls), 5),
             sd_ll   = round(sd(lls), 5),
             min_ll  = round(min(lls), 5),
             max_ll  = round(max(lls), 5))
}

comparison <- do.call(rbind, lapply(variants, summarise_variant))
comparison <- comparison[order(comparison$mean_ll), ]
comparison
```

# Save the comparison (one row per variant)

```{r}
path <- "mnl_improvement_stats.csv"
out  <- comparison; out$date <- as.character(Sys.Date())
if (file.exists(path)) {
  write.csv(rbind(read.csv(path, stringsAsFactors=FALSE), out), path, row.names=FALSE)
} else write.csv(out, path, row.names=FALSE)
cat("Saved to", path, "\n")
```

# How to read it

- **mean_ll** — expected score (lower better).
- **sd_ll** — stability across splits (lower safer for the private board).
- Keep a variant ONLY if it lowers mean_ll without meaningfully raising sd_ll
  vs baseline. A tiny mean gain with a big sd jump is not worth it.
- The winner on BOTH: refit on all train and generate the submission
  (final_submission_mnl.Rmd, swapping in the winning formula/columns).
```
