---
title: "draft"
output: html_document
date: "2026-07-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Inspect data

```{r}
train <- read.csv("data/train.csv")
test  <- read.csv("data/test.csv")

dim(train)   # 21565 x 113
dim(test)    # 4997  x 113

str(train)
str(test)
```

# Summary

```{r}
summary(train)
head(train)
```

# Derive the single Choice variable

```{r}
train$Choice <- max.col(train[, c("Ch1","Ch2","Ch3","Ch4")])
table(train$Choice)

which(colnames(train)=="CC1")   # 4
which(colnames(train)=="Price4") # 83
```

# Reshape training data (Task <= 12) for mlogit

```{r}
library(mlogit)
S <- dfidx(subset(train, Task<=12), shape="wide", choice="Choice",
           sep="", varying = c(4:83), idx = c("No", "Case"))
head(S)
```

# Fit MNL + training log loss

```{r}
M <- mlogit(Choice~CC+GN+NS+BU+FA+LD+BZ+FC+FP+RP+PP+KA+SC+TS+NV+MA+LB+AF+HU+Price-1, data=S)
summary(M)

# --- log loss (competition metric) ---
log_loss <- function(actual, pred, eps = 1e-15) {
  actual <- as.matrix(actual)
  pred   <- as.matrix(pred)
  stopifnot(all(dim(actual) == dim(pred)))
  pred <- pred / rowSums(pred)                 # rescale to sum to 1
  pred <- pmin(pmax(pred, eps), 1 - eps)        # clip for numerical safety
  -mean(rowSums(actual * log(pred)))
}

ActualChoice <- subset(train, Task<=12)[,"Choice"]
P <- predict(M, newdata=S)

truth_mat <- matrix(0, length(ActualChoice), 4)
truth_mat[cbind(seq_along(ActualChoice), ActualChoice)] <- 1

cat("Training log loss:", round(log_loss(truth_mat, P), 5), "\n")

# benchmark sanity check (should be ~1.38629)
bench <- matrix(0.25, length(ActualChoice), 4)
cat("Benchmark (0.25 each):", round(log_loss(truth_mat, bench), 5), "\n")

# accuracy table
PredictedChoice <- apply(P, 1, which.max)
Tabtrain <- table(PredictedChoice, ActualChoice)
Tabtrain
cat("Training accuracy:", round(sum(diag(Tabtrain))/sum(Tabtrain), 4), "\n")
```

# Validation on held-out tasks (Task > 12)

This is still train.csv data — a held-out slice we can score, NOT the
competition test set.

```{r}
Val <- dfidx(subset(train, Task > 12), shape = "wide", choice = "Choice",
             sep = "", varying = c(4:83), idx = c("No", "Case"))

ValPredict   <- predict(M, newdata = Val)          # n x 4 probability matrix
ActualChoiceV <- subset(train, Task > 12)[, "Choice"]

# accuracy
PredictedChoiceV <- apply(ValPredict, 1, which.max)
Tabval <- table(PredictedChoiceV, ActualChoiceV)
Tabval
cat("Validation accuracy:", round(sum(diag(Tabval)) / sum(Tabval), 4), "\n")

# validation log loss (our real held-out estimate)
truth_matV <- matrix(0, length(ActualChoiceV), 4)
truth_matV[cbind(seq_along(ActualChoiceV), ActualChoiceV)] <- 1
cat("Validation log loss:", round(log_loss(truth_matV, ValPredict), 5), "\n")
```

# GAP SOLVED: refit on ALL of train, predict on the real test.csv, submit

Now we use the actual competition test set. `test.csv` has blank `Ch1..Ch4`,
so we add a placeholder `Choice` just so `dfidx` can build the object; the
values are never used for prediction.

```{r}
# placeholder choice so dfidx works on the unlabelled test set
test$Choice <- 1L

# refit on ALL 19 tasks of train (not just Task<=12) so we use every observation
S_all  <- dfidx(train, shape="wide", choice="Choice",
                sep="", varying = c(4:83), idx = c("No", "Case"))
S_test <- dfidx(test,  shape="wide", choice="Choice",
                sep="", varying = c(4:83), idx = c("No", "Case"))

M_full <- mlogit(Choice~CC+GN+NS+BU+FA+LD+BZ+FC+FP+RP+PP+KA+SC+TS+NV+MA+LB+AF+HU+Price-1,
                 data = S_all)

test_prob <- predict(M_full, newdata = S_test)   # 4997 x 4 probabilities

# build submission in the exact sample format: No,Ch1,Ch2,Ch3,Ch4
sample_sub <- read.csv("data/sample_submission.csv")
submission <- data.frame(No = test$No,
                         Ch1 = test_prob[,1], Ch2 = test_prob[,2],
                         Ch3 = test_prob[,3], Ch4 = test_prob[,4])

# safety checks
stopifnot(all(names(submission) == names(sample_sub)))   # columns match
stopifnot(all(submission$No == sample_sub$No))           # ids align
cat("row sums:", range(rowSums(submission[,-1])), " NA:", anyNA(submission), "\n")

if (!dir.exists("submissions")) dir.create("submissions")
write.csv(submission, "submissions/submission_mnl.csv", row.names = FALSE)
cat("Wrote submissions/submission_mnl.csv — ready to upload to Kaggle\n")
```
