---
title: "Final submission - MNL with selective factor attributes"
output: html_document
date: "2026-07-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Refit the winning "factors" MNL on ALL training data, predict the test set, and
write the Kaggle submission. Respondent-level stability run gave
**mean 1.20174, sd 0.01527** → expected leaderboard roughly **1.187–1.217**
(mean +/- 1 sd).

Caution handled: factor levels (and rare-level collapsing) are learned on the
FULL training data and applied to test with the SAME levels, so there is no
level-mismatch error.

# Load

```{r}
library(mlogit)
train <- read.csv("/Users/imeldalee/Desktop/TAE KAGGLE/kagglecomp/data/train.csv")
test  <- read.csv("/Users/imeldalee/Desktop/TAE KAGGLE/kagglecomp/data/test.csv")

train$Choice <- max.col(train[, c("Ch1","Ch2","Ch3","Ch4")])
test$Choice  <- 1L

attrs    <- c("CC","GN","NS","BU","FA","LD","BZ","FC","FP","RP",
              "PP","KA","SC","TS","NV","MA","LB","AF","HU")
attr_rhs <- paste(c(attrs, "Price"), collapse = "+")

ftr <- c("NS","BU","FP","SC","MA","LB")   # the 6 factored attributes
```

# Collapse rare levels — learned on TRAIN, applied to TEST identically

```{r}
# For each factored attribute's 4 alternative columns, find the levels that are
# common enough in TRAIN (>= 30), and the modal level. Store them so we apply
# the EXACT same mapping to test (this is the caution being handled).
level_map <- list()
for (a in ftr) {
  # pool all 4 alternative columns of this attribute across train to decide levels
  allvals <- unlist(train[, paste0(a, 1:4)])
  tab   <- table(allvals)
  keep  <- as.numeric(names(tab)[tab >= 30])
  modal <- as.numeric(names(which.max(tab)))
  level_map[[a]] <- list(keep = keep, modal = modal)
}

collapse_rare <- function(df) {
  for (a in ftr) {
    keep  <- level_map[[a]]$keep
    modal <- level_map[[a]]$modal
    for (j in 1:4) {
      col <- paste0(a, j)
      df[[col]][!(df[[col]] %in% keep)] <- modal
    }
  }
  df
}

train <- collapse_rare(train)
test  <- collapse_rare(test)   # SAME keep/modal from train
```

# Refit the factors MNL on ALL training data

```{r}
S_all  <- dfidx(train, shape="wide", choice="Choice", sep="",
                varying = c(4:83), idx = c("No","Case"))
S_test <- dfidx(test,  shape="wide", choice="Choice", sep="",
                varying = c(4:83), idx = c("No","Case"))

# factor the 6 attributes; store train levels and apply the SAME to test
for (a in ftr) {
  S_all[[a]]  <- factor(S_all[[a]])
  S_test[[a]] <- factor(S_test[[a]], levels = levels(S_all[[a]]))  # <- alignment
}

M_full <- mlogit(as.formula(paste("Choice ~", attr_rhs, "- 1")), data = S_all)
summary(M_full)
```

# Predict test

```{r}
test_prob <- predict(M_full, newdata = S_test)
dim(test_prob)                 # 4997 x 4
summary(rowSums(test_prob))    # all 1
cat("any NA in predictions:", anyNA(test_prob), "\n")
```

# Build, check, write submission

```{r}
sample_sub <- read.csv("/Users/imeldalee/Desktop/TAE KAGGLE/kagglecomp/data/sample_submission.csv")
submission <- data.frame(No = test$No,
                         Ch1 = test_prob[,1], Ch2 = test_prob[,2],
                         Ch3 = test_prob[,3], Ch4 = test_prob[,4])

stopifnot(all(names(submission) == names(sample_sub)))
stopifnot(all(submission$No == sample_sub$No))
stopifnot(!anyNA(submission))
cat("rows:", nrow(submission), " row-sum range:", range(rowSums(submission[,-1])), "\n")

if (!dir.exists("submissions")) dir.create("submissions")
write.csv(submission, "submissions/submission_factors.csv", row.names = FALSE)
cat("Wrote submissions/submission_factors.csv\n")
```

# Log the submission (with the stability range)

```{r}
mean_ll <- 1.20174
sd_ll   <- 0.01527
range_txt <- sprintf("expected ~%.3f (sd %.3f; ~%.3f-%.3f)",
                     mean_ll, sd_ll, mean_ll - sd_ll, mean_ll + sd_ll)

path <- "/Users/imeldalee/Desktop/TAE KAGGLE/kagglecomp/submissions_log.csv"
log  <- read.csv(path, stringsAsFactors = FALSE)
new  <- data.frame(
  date              = as.character(Sys.Date()),
  model_name        = "mnl_factors",
  description       = "MNL with selective factor attributes (NS,BU,FP,SC,MA,LB; rare levels<30 collapsed to modal), refit on all train",
  source_citation   = "Class material (mlogit / dfidx)",
  cv_logloss        = mean_ll,
  public_lb_logloss = NA, gap = NA,
  notes             = paste("Respondent-level 20-split stability:", range_txt,
                            "| levels learned on train, applied to test."),
  stringsAsFactors  = FALSE
)
write.csv(rbind(log, new), path, row.names = FALSE)
cat("Logged mnl_factors |", range_txt, "\n")
```

# After Kaggle: record the public score and see if it lands in range

```{r eval=FALSE}
log <- read.csv("/Users/imeldalee/Desktop/TAE KAGGLE/kagglecomp/submissions_log.csv", stringsAsFactors = FALSE)
i <- which(log$model_name == "mnl_factors")[1]
log$public_lb_logloss[i] <- 0.000   # <- Kaggle public score here
log$gap[i] <- log$public_lb_logloss[i] - log$cv_logloss[i]
write.csv(log, "/Users/imeldalee/Desktop/TAE KAGGLE/kagglecomp/submissions_log.csv", row.names = FALSE)
cat("public:", log$public_lb_logloss[i], "| expected ~1.187-1.217\n")
```
