---
title: "Model families 1-3: MNL, Mixed Logit, Hierarchical Bayes"
output: html_document
date: "2026-07-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Experiments with three choice-model families on the same respondent-level
validation split, scored by log-loss. Same simple `dfidx`/`mlogit` setup as
`draft.Rmd`.

# Load & prepare

```{r}
train <- read.csv("/Users/imeldalee/Desktop/TAE KAGGLE/kagglecomp/data/train.csv")
test  <- read.csv("/Users/imeldalee/Desktop/TAE KAGGLE/kagglecomp/data/test.csv")

train$Choice <- max.col(train[, c("Ch1","Ch2","Ch3","Ch4")])   # 1..4
test$Choice  <- 1L                                             # placeholder

which(colnames(train)=="CC1")    # 4
which(colnames(train)=="Price4") # 83
```

# Log-loss helper

```{r}
log_loss <- function(actual, pred, eps = 1e-15) {
  actual <- as.matrix(actual); pred <- as.matrix(pred)
  stopifnot(all(dim(actual) == dim(pred)))
  pred <- pred / rowSums(pred)
  pred <- pmin(pmax(pred, eps), 1 - eps)
  -mean(rowSums(actual * log(pred)))
}

# turn a 1..4 vector into a one-hot n x 4 matrix
one_hot <- function(v) {
  m <- matrix(0, length(v), 4)
  m[cbind(seq_along(v), v)] <- 1
  m
}
```

# Split: train on Task<=12, validate on Task>12 (class-style)

```{r}
library(mlogit)

train_fit <- subset(train, Task <= 12)
train_val <- subset(train, Task > 12)

val_truth <- one_hot(train_val$Choice)

# shared reshape helpers
S_fit  <- dfidx(train_fit, shape="wide", choice="Choice", sep="",
                varying = c(4:83), idx = c("No","Case"))
S_val  <- dfidx(train_val, shape="wide", choice="Choice", sep="",
                varying = c(4:83), idx = c("No","Case"))

# panel versions (list idx) for mixed logit
S_fitP <- dfidx(train_fit, shape="wide", choice="Choice", sep="",
                varying = c(4:83), idx = list(c("No","Case")))
S_valP <- dfidx(train_val, shape="wide", choice="Choice", sep="",
                varying = c(4:83), idx = list(c("No","Case")))

attr_rhs <- "CC+GN+NS+BU+FA+LD+BZ+FC+FP+RP+PP+KA+SC+TS+NV+MA+LB+AF+HU+Price"
```

# Family 1 — Conditional / Multinomial Logit (baseline)

```{r}
M1 <- mlogit(as.formula(paste("Choice ~", attr_rhs, "- 1")), data = S_fit)
summary(M1)

p1  <- predict(M1, newdata = S_val)
ll1 <- log_loss(val_truth, p1)
cat("MNL validation log loss:", round(ll1, 5), "\n")
```

# Family 2 — Mixed (random-parameters) Logit

Random-normal coefficients across respondents, panel over `Case`.
Start with a random Price coefficient (fast); the commented block makes all
coefficients random (slower, matches the class example).

```{r}
rpar_price <- c(Price = "n")

M2 <- mlogit(as.formula(paste("Choice ~", attr_rhs, "- 1")), data = S_fitP,
             rpar = rpar_price, panel = TRUE, R = 50, halton = NA)
summary(M2)

p2  <- predict(M2, newdata = S_valP)
ll2 <- log_loss(val_truth, p2)
cat("Mixed logit (random Price) validation log loss:", round(ll2, 5), "\n")
```

```{r eval=FALSE}
# ALL coefficients random (slower; raise R for final fit)
attr_vars <- c("CC","GN","NS","BU","FA","LD","BZ","FC","FP","RP",
               "PP","KA","SC","TS","NV","MA","LB","AF","HU","Price")
rpar_all  <- setNames(rep("n", length(attr_vars)), attr_vars)
M2b <- mlogit(as.formula(paste("Choice ~", attr_rhs, "- 1")), data = S_fitP,
              rpar = rpar_all, panel = TRUE, R = 50, halton = NA)
p2b <- predict(M2b, newdata = S_valP)
cat("Mixed logit (all random) val log loss:", round(log_loss(val_truth, p2b),5), "\n")
```

# Family 3 — Hierarchical Bayes MNL

Uses `ChoiceModelR` (wrapper on bayesm's hierarchical MNL sampler). HB gives
individual-level partworths; holdout probabilities are averaged over posterior
draws. This is slower (MCMC) and needs a specific long-format input.

```{r eval=FALSE}
# install.packages("ChoiceModelR")
library(ChoiceModelR)

# ChoiceModelR wants long format:
#   columns: id (respondent), set (task), alt (1..4), then attributes, then choice
# Build it from the wide train_fit.
attrs <- c("CC","GN","NS","BU","FA","LD","BZ","FC","FP","RP",
           "PP","KA","SC","TS","NV","MA","LB","AF","HU")
build_hb_long <- function(df) {
  do.call(rbind, lapply(1:4, function(j) {
    out <- data.frame(id = df$Case, set = df$Task, alt = j)
    for (a in attrs) out[[a]] <- df[[paste0(a, j)]]
    out$Price <- df[[paste0("Price", j)]]
    # choice: 1..4 on the alt==1 row only (ChoiceModelR convention:
    # the chosen alt number goes in the 'y' column, 0 for non-first rows)
    out$y <- ifelse(j == 1, df$Choice, 0)
    out
  }))
}
hb_long <- build_hb_long(train_fit)
hb_long <- hb_long[order(hb_long$id, hb_long$set, hb_long$alt), ]

xcoding <- rep(1, length(attrs) + 1)   # 1 = treat as continuous/linear
mcmc  <- list(R = 2000, use = 1000)    # draws; use last 1000
options <- list(none = FALSE, save = TRUE)

hb_out <- choicemodelr(hb_long[, c("id","set","alt", attrs, "Price","y")],
                       xcoding, mcmc = mcmc, options = options)
# hb_out$betadraw holds per-respondent posterior draws; average choice
# probabilities over draws for holdout prediction.
```

> Note: HB in R (`ChoiceModelR` / `bayesm::rhierMnlRwMixture`) needs careful
> long-format setup and MCMC time. If it's fiddly, the mixed logit (Family 2)
> gives near-identical predictive performance with a simpler `mlogit` call — so
> treat HB as optional / for the report's discussion of heterogeneity.

# Comparison

```{r}
compare <- data.frame(
  model = c("1_MNL", "2_mixed_logit_price"),
  val_logloss = round(c(ll1, ll2), 5)
)
compare <- compare[order(compare$val_logloss), ]
compare

# benchmark check
bench <- matrix(0.25, nrow(val_truth), 4)
cat("benchmark (0.25 each):", round(log_loss(val_truth, bench), 5), "\n")
```

# Notes

- MNL and mixed logit both run via `mlogit`; HB (Family 3) is heavier and
  optional — the research suggests HB and mixed logit are near-equivalent
  predictively, so Family 2 is the practical "heterogeneity" model.
- To generate a submission from the winner: refit on ALL of train
  (all 19 tasks), predict on `test`, write `No,Ch1,Ch2,Ch3,Ch4` (see draft.Rmd
  submission chunk).
```
