---
title: "04 - Final Model & Submission"
author: "Team"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
source(here("R", "log_loss.R"))
```

> This notebook holds the **final chosen model** and the code that generates
> the submission we actually send to Kaggle. Keep it clean — the report is
> written from what's here. Experiments live in `03_model_experiments.Rmd`
> and `experiments/`.

# Load data + validation split

```{r load}
train <- read.csv(here("data", "train.csv"))
test  <- read.csv(here("data", "test.csv"))
# split <- readRDS(here("data_processed", "train_val_split.rds"))
```

# Final model specification

Describe in one or two sentences *why* this is the final model (best
validation log loss, robust public LB, interpretable, etc.).

```{r final-model}
# Fit the final model here.
# final_fit <- ...
# saveRDS(final_fit, here("models", "final_model.rds"))
```

# Validation performance

```{r validation}
# pred_val <- predict(final_fit, ...)
# log_loss(pred_val, truth_val)
```

# Refit on ALL training data before predicting test

Once the model is chosen, refit on the full training set (train + validation)
so the final predictions use every available observation.

```{r refit-full}
# final_fit_full <- ... (same spec, all of train.csv)
```

# Predict on test and build submission

```{r predict-test}
# pred_test <- predict(final_fit_full, newdata = test, ...)

# Match sample_submission.csv format EXACTLY (column names + order)
# sample_sub <- read.csv(here("data", "sample_submission.csv"))
# names(sample_sub)

# Rows must sum to 1 across the 4 alternatives (rescale if needed)
# sub <- data.frame(...)
# write.csv(sub, here("submissions", "submission_final.csv"), row.names = FALSE)
```

# Final submission checklist

- [ ] Columns and order match `sample_submission.csv`
- [ ] Each row's 4 probabilities sum to 1
- [ ] No NAs in predictions
- [ ] Row count matches test.csv (4997)
- [ ] Logged in `submissions_log.csv` (val score, public LB, source citation)
- [ ] Model object saved to `models/`
