---
title: "Blend weight sweep - find the best pull toward uniform"
output: html_document
date: "2026-07-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Blending toward 0.25 was the winning robustness knob. Here we sweep the blend
weight (lambda) over 100 respondent-level seeds to find the value that minimizes
mean log-loss while keeping the sd (70/30 stability) low. Uses the full
covariate-interaction MNL.

# Setup (same pieces as robustness_pass)

```{r}
library(mlogit)
train <- read.csv("data/train.csv")
train$Choice <- max.col(train[, c("Ch1","Ch2","Ch3","Ch4")])

attrs <- c("CC","GN","NS","BU","FA","LD","BZ","FC","FP","RP",
           "PP","KA","SC","TS","NV","MA","LB","AF","HU")
ftr   <- c("NS","BU","FP","SC","MA","LB")

log_loss <- function(actual, pred, eps=1e-15){actual<-as.matrix(actual);pred<-as.matrix(pred)
  pred<-pred/rowSums(pred);pred<-pmin(pmax(pred,eps),1-eps);-mean(rowSums(actual*log(pred)))}
one_hot <- function(v){m<-matrix(0,length(v),4);m[cbind(seq_along(v),v)]<-1;m}
blend_prior <- function(P, lambda){(1-lambda)*P + lambda*0.25}

scale_cov <- function(df){df$incomea_s<-df$incomea/1e5;df$agea_s<-df$agea/50
  df$milesa_s<-df$milesa/100;df$nighta_s<-df$nighta/50;df}
add_interactions <- function(df){df<-scale_cov(df)
  for(j in 1:4){df[[paste0("PxINC",j)]]<-df[[paste0("Price",j)]]*df$incomea_s
    df[[paste0("PxSEG",j)]]<-df[[paste0("Price",j)]]*df$segmentind
    df[[paste0("PxAGE",j)]]<-df[[paste0("Price",j)]]*df$agea_s
    df[[paste0("PxMIL",j)]]<-df[[paste0("Price",j)]]*df$milesa_s
    df[[paste0("PxNGT",j)]]<-df[[paste0("Price",j)]]*df$nighta_s}; df}
stems <- c("PxINC","PxSEG","PxAGE","PxMIL","PxNGT")
```

# Cache the raw predictions once per split (so we don't refit for each lambda)

```{r}
seeds <- 1:100
cache <- lapply(seeds, function(seed){
  set.seed(seed)
  vc<-sample(unique(train$Case),round(0.20*length(unique(train$Case))))
  fit<-subset(train,!(Case %in% vc)); val<-subset(train,Case %in% vc)
  for(a in ftr) for(j in 1:4){col<-paste0(a,j);tb<-table(fit[[col]])
    keep<-as.numeric(names(tb)[tb>=30]);modal<-as.numeric(names(which.max(tb)))
    fit[[col]][!(fit[[col]]%in%keep)]<-modal; val[[col]][!(val[[col]]%in%keep)]<-modal}
  fit<-add_interactions(fit); val<-add_interactions(val)
  extra<-unlist(lapply(stems,function(s)paste0(s,1:4)))
  vn<-c(names(fit)[4:83],extra)
  Sf<-dfidx(fit,shape="wide",choice="Choice",sep="",varying=which(names(fit)%in%vn),idx=c("No","Case"))
  Sv<-dfidx(val,shape="wide",choice="Choice",sep="",varying=which(names(val)%in%vn),idx=c("No","Case"))
  for(a in ftr){Sf[[a]]<-factor(Sf[[a]]);Sv[[a]]<-factor(Sv[[a]],levels=levels(Sf[[a]]))}
  rhs<-paste(paste(c(attrs,"Price"),collapse="+"),"+",paste(stems,collapse="+"))
  M<-mlogit(as.formula(paste("Choice ~",rhs,"- 1")),data=Sf)
  list(P=predict(M,newdata=Sv), truth=one_hot(val$Choice))
})
cat("cached", length(cache), "splits\n")
```

# Sweep lambda over the cached predictions

```{r}
batches <- split(seq_along(seeds), ceiling(seq_along(seeds)/20))
lambdas <- c(0, 0.02, 0.03, 0.05, 0.08, 0.10, 0.12, 0.15, 0.20)

sweep <- do.call(rbind, lapply(lambdas, function(lam){
  all_ll<-c(); bmeans<-c()
  for(b in batches){
    lls <- sapply(b, function(i) log_loss(cache[[i]]$truth, blend_prior(cache[[i]]$P, lam)))
    all_ll<-c(all_ll,lls); bmeans<-c(bmeans,mean(lls))
  }
  data.frame(lambda=lam, mean_ll=round(mean(all_ll),5),
             sd=round(sd(all_ll),5), se=round(sd(bmeans),5))
}))
sweep <- sweep[order(sweep$mean_ll), ]
sweep
```

# Best lambda

```{r}
best <- sweep$lambda[which.min(sweep$mean_ll)]
cat("Best blend lambda by mean:", best, "\n")
cat("(If several lambdas tie on mean, prefer the larger one - more shrinkage,\n")
cat(" lower sd, safer on the hidden 30%.)\n")
plot(sweep$lambda[order(sweep$lambda)],
     sweep$mean_ll[order(sweep$lambda)], type="b", pch=19,
     xlab="blend lambda (pull toward 0.25)", ylab="mean log loss (100 seeds)",
     main="Blend weight sweep")
write.csv(sweep, "blend_sweep.csv", row.names=FALSE)
```

# Generate the final blended submission with the chosen lambda

```{r eval=FALSE}
LAMBDA <- best   # or set manually, e.g. 0.08

test <- read.csv("data/test.csv"); test$Choice <- 1L
level_map<-list()
for(a in ftr){allv<-unlist(train[,paste0(a,1:4)]);tb<-table(allv)
  level_map[[a]]<-list(keep=as.numeric(names(tb)[tb>=30]),modal=as.numeric(names(which.max(tb))))}
collapse_rare<-function(df){for(a in ftr)for(j in 1:4){col<-paste0(a,j)
  df[[col]][!(df[[col]]%in%level_map[[a]]$keep)]<-level_map[[a]]$modal};df}

trX<-add_interactions(collapse_rare(train)); teX<-add_interactions(collapse_rare(test))
extra<-unlist(lapply(stems,function(s)paste0(s,1:4))); vn<-c(names(trX)[4:83],extra)
S_all <-dfidx(trX,shape="wide",choice="Choice",sep="",varying=which(names(trX)%in%vn),idx=c("No","Case"))
S_test<-dfidx(teX,shape="wide",choice="Choice",sep="",varying=which(names(teX)%in%vn),idx=c("No","Case"))
for(a in ftr){S_all[[a]]<-factor(S_all[[a]]);S_test[[a]]<-factor(S_test[[a]],levels=levels(S_all[[a]]))}
rhs<-paste(paste(c(attrs,"Price"),collapse="+"),"+",paste(stems,collapse="+"))
M<-mlogit(as.formula(paste("Choice ~",rhs,"- 1")),data=S_all)
P<-blend_prior(predict(M,newdata=S_test), LAMBDA); P<-P/rowSums(P)

sub<-data.frame(No=test$No,Ch1=P[,1],Ch2=P[,2],Ch3=P[,3],Ch4=P[,4])
sample_sub<-read.csv("data/sample_submission.csv")
stopifnot(all(names(sub)==names(sample_sub)),all(sub$No==sample_sub$No),!anyNA(sub))
write.csv(sub,"submissions/submission_covar_blend.csv",row.names=FALSE)

log<-read.csv("submissions_log.csv",stringsAsFactors=FALSE)
new<-data.frame(date=as.character(Sys.Date()),model_name="mnl_covar_blend",
  description=sprintf("Covariate MNL, predictions blended %.2f toward 0.25 uniform",LAMBDA),
  source_citation="own experiment",
  cv_logloss=sweep$mean_ll[sweep$lambda==LAMBDA][1],public_lb_logloss=NA,gap=NA,
  notes="Blend chosen by 100-seed sweep for lowest mean + stable sd.",stringsAsFactors=FALSE)
write.csv(rbind(log,new),"submissions_log.csv",row.names=FALSE)
cat("Wrote submissions/submission_covar_blend.csv with lambda",LAMBDA,"\n")
```
```
