---
title: "Factors MNL + respondent-covariate interactions"
output: html_document
date: "2026-07-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The change teammate feedback flagged as the biggest generalizable win: interact
respondent covariates (income, age, miles, night, segment) with **price** (and
optionally key attributes), so different respondent types weigh price/features
differently. This is the main lever for predicting UNSEEN respondents, since it
lets a new person's covariates shift their predicted choices.

Built as `Price * covariate` columns (alt-varying: 0 for the opt-out since
Price4=0), so mlogit treats them as ordinary generic-coefficient terms.
Validated on the same 20 respondent-level splits.

# Setup

```{r}
library(mlogit)
train <- read.csv("/Users/imeldalee/Desktop/TAE KAGGLE/kagglecomp/data/train.csv")
test  <- read.csv("/Users/imeldalee/Desktop/TAE KAGGLE/kagglecomp/data/test.csv")
train$Choice <- max.col(train[, c("Ch1","Ch2","Ch3","Ch4")])
test$Choice  <- 1L

attrs <- c("CC","GN","NS","BU","FA","LD","BZ","FC","FP","RP",
           "PP","KA","SC","TS","NV","MA","LB","AF","HU")

# covariates to interact with price (numeric versions; scaled so coeffs behave)
covars_num <- c(incomea="incomea", agea="agea", milesa="milesa", nighta="nighta")
# categorical covariate codes to interact (kept as-is, small integer codes)
covars_cat <- c(segmentind="segmentind")

ftr <- c("NS","BU","FP","SC","MA","LB")   # factored attributes

log_loss <- function(actual, pred, eps = 1e-15) {
  actual <- as.matrix(actual); pred <- as.matrix(pred)
  stopifnot(all(dim(actual) == dim(pred)))
  pred <- pred / rowSums(pred); pred <- pmin(pmax(pred, eps), 1 - eps)
  -mean(rowSums(actual * log(pred)))
}
one_hot <- function(v) { m <- matrix(0, length(v), 4); m[cbind(seq_along(v), v)] <- 1; m }
```

# Build Price x covariate interaction columns (alt-varying, no separator)

```{r}
# scale numeric covariates to ~unit range so interaction coefficients are stable
scale_cov <- function(df) {
  df$incomea_s <- df$incomea / 1e5
  df$agea_s    <- df$agea    / 50
  df$milesa_s  <- df$milesa  / 100
  df$nighta_s  <- df$nighta  / 50
  df
}

add_interactions <- function(df) {
  df <- scale_cov(df)
  # Price x each scaled numeric covariate, per alternative (PxINC1..4, etc.)
  for (j in 1:4) {
    df[[paste0("PxINC", j)]] <- df[[paste0("Price", j)]] * df$incomea_s
    df[[paste0("PxAGE", j)]] <- df[[paste0("Price", j)]] * df$agea_s
    df[[paste0("PxMIL", j)]] <- df[[paste0("Price", j)]] * df$milesa_s
    df[[paste0("PxNGT", j)]] <- df[[paste0("Price", j)]] * df$nighta_s
    df[[paste0("PxSEG", j)]] <- df[[paste0("Price", j)]] * df$segmentind
  }
  df
}

inter_stems <- c("PxINC","PxAGE","PxMIL","PxNGT","PxSEG")
```

# One split -> fit -> validation log-loss

```{r}
run_cov <- function(seed, use_interactions = TRUE) {
  set.seed(seed)
  vc  <- sample(unique(train$Case), round(0.20*length(unique(train$Case))))
  fit <- subset(train, !(Case %in% vc)); val <- subset(train, Case %in% vc)
  truth <- one_hot(val$Choice)

  # collapse rare levels for the factored attrs (train-derived)
  for (a in ftr) for (j in 1:4) {
    col <- paste0(a,j); tab <- table(fit[[col]])
    keep <- as.numeric(names(tab)[tab>=30]); modal <- as.numeric(names(which.max(tab)))
    fit[[col]][!(fit[[col]] %in% keep)] <- modal
    val[[col]][!(val[[col]] %in% keep)] <- modal
  }

  fit <- add_interactions(fit); val <- add_interactions(val)

  # varying columns: base attrs+price (4:83) plus the interaction blocks
  extra <- unlist(lapply(inter_stems, function(s) paste0(s, 1:4)))
  vary_names <- c(names(fit)[4:83], if (use_interactions) extra else character(0))

  Sf <- dfidx(fit, shape="wide", choice="Choice", sep="",
              varying = which(names(fit) %in% vary_names), idx=c("No","Case"))
  Sv <- dfidx(val, shape="wide", choice="Choice", sep="",
              varying = which(names(val) %in% vary_names), idx=c("No","Case"))
  for (a in ftr) { Sf[[a]] <- factor(Sf[[a]]); Sv[[a]] <- factor(Sv[[a]], levels=levels(Sf[[a]])) }

  rhs <- paste(c(attrs, "Price"), collapse="+")
  if (use_interactions) rhs <- paste(rhs, "+", paste(inter_stems, collapse="+"))

  M <- mlogit(as.formula(paste("Choice ~", rhs, "- 1")), data = Sf)
  p <- predict(M, newdata = Sv)
  log_loss(truth, p)
}
```

# Stability: factors-only vs factors + covariate interactions

```{r}
seeds <- 1:20
summ <- function(fn) {
  x <- sapply(seeds, function(s) tryCatch(fn(s), error=function(e){message(e);NA}))
  x <- x[!is.na(x)]
  c(n=length(x), mean=round(mean(x),5), sd=round(sd(x),5),
    min=round(min(x),5), max=round(max(x),5))
}

no_int  <- summ(function(s) run_cov(s, use_interactions = FALSE))
with_int<- summ(function(s) run_cov(s, use_interactions = TRUE))

rbind(factors_only = no_int, factors_plus_covar_x_price = with_int)
```

# Robustness check: does blending toward uniform reduce error/variance?

From the earlier robustness pass + 100-seed sweep, blending predictions gently
toward the 0.25 uniform prior lowered the spread (sd) across splits without
hurting the mean — making the single submission more stable across whatever
70/30 slice Kaggle draws. lambda=0.08 was the best mean+stability trade-off.
Here we re-confirm it on the covariate model's own splits.

```{r}
blend_prior <- function(P, lambda){ P <- (1-lambda)*P + lambda*0.25; P/rowSums(P) }

# run the covariate model once per split, score raw vs blended at a few lambdas
seeds_r <- 1:20
lambdas <- c(0, 0.05, 0.08, 0.10)

# cache raw predictions + truth per split (refit once, blend cheaply)
cache <- lapply(seeds_r, function(s){
  set.seed(s)
  vc<-sample(unique(train$Case),round(0.20*length(unique(train$Case))))
  fit<-subset(train,!(Case %in% vc)); val<-subset(train,Case %in% vc)
  for(a in ftr) for(j in 1:4){col<-paste0(a,j);tb<-table(fit[[col]])
    keep<-as.numeric(names(tb)[tb>=30]);modal<-as.numeric(names(which.max(tb)))
    fit[[col]][!(fit[[col]]%in%keep)]<-modal; val[[col]][!(val[[col]]%in%keep)]<-modal}
  fit<-add_interactions(fit); val<-add_interactions(val)
  extra<-unlist(lapply(inter_stems,function(x)paste0(x,1:4)))
  vn<-c(names(fit)[4:83],extra)
  Sf<-dfidx(fit,shape="wide",choice="Choice",sep="",varying=which(names(fit)%in%vn),idx=c("No","Case"))
  Sv<-dfidx(val,shape="wide",choice="Choice",sep="",varying=which(names(val)%in%vn),idx=c("No","Case"))
  for(a in ftr){Sf[[a]]<-factor(Sf[[a]]);Sv[[a]]<-factor(Sv[[a]],levels=levels(Sf[[a]]))}
  rhs<-paste(paste(c(attrs,"Price"),collapse="+"),"+",paste(inter_stems,collapse="+"))
  M<-mlogit(as.formula(paste("Choice ~",rhs,"- 1")),data=Sf)
  list(P=predict(M,newdata=Sv), truth=one_hot(val$Choice))
})

blend_tbl <- do.call(rbind, lapply(lambdas, function(lam){
  lls <- sapply(cache, function(c) log_loss(c$truth, blend_prior(c$P, lam)))
  data.frame(lambda=lam, mean_ll=round(mean(lls),5), sd_ll=round(sd(lls),5))
}))
blend_tbl
cat("\nPick lambda with low mean AND low sd. Earlier full 100-seed sweep chose 0.08.\n")
BEST_LAMBDA <- 0.08
```

# Refit on all train, predict test, apply blend, write + log

```{r}
# level alignment for factors (train-derived, applied to test)
level_map <- list()
for (a in ftr) { allv <- unlist(train[,paste0(a,1:4)]); tb <- table(allv)
  level_map[[a]] <- list(keep=as.numeric(names(tb)[tb>=30]), modal=as.numeric(names(which.max(tb)))) }
collapse_rare <- function(df) { for (a in ftr) for (j in 1:4) {
  col<-paste0(a,j); df[[col]][!(df[[col]] %in% level_map[[a]]$keep)] <- level_map[[a]]$modal }; df }

trX <- add_interactions(collapse_rare(train))
teX <- add_interactions(collapse_rare(test))

extra <- unlist(lapply(inter_stems, function(s) paste0(s, 1:4)))
vary_names <- c(names(trX)[4:83], extra)
S_all  <- dfidx(trX, shape="wide", choice="Choice", sep="",
                varying=which(names(trX) %in% vary_names), idx=c("No","Case"))
S_test <- dfidx(teX, shape="wide", choice="Choice", sep="",
                varying=which(names(teX) %in% vary_names), idx=c("No","Case"))
for (a in ftr) { S_all[[a]] <- factor(S_all[[a]]); S_test[[a]] <- factor(S_test[[a]], levels=levels(S_all[[a]])) }

rhs <- paste(paste(c(attrs,"Price"),collapse="+"), "+", paste(inter_stems, collapse="+"))
Mfull <- mlogit(as.formula(paste("Choice ~", rhs, "- 1")), data = S_all)
P_raw <- predict(Mfull, newdata = S_test)

# ROBUSTNESS: blend toward uniform (lambda from the sweep), which also floors
# every probability away from 0 -> caps worst-case log-loss spikes on unlucky rows
P <- blend_prior(P_raw, BEST_LAMBDA)

sample_sub <- read.csv("/Users/imeldalee/Desktop/TAE KAGGLE/kagglecomp/data/sample_submission.csv")
submission <- data.frame(No=test$No, Ch1=P[,1], Ch2=P[,2], Ch3=P[,3], Ch4=P[,4])
stopifnot(all(names(submission)==names(sample_sub)), all(submission$No==sample_sub$No), !anyNA(submission))
cat("row-sum range:", range(rowSums(submission[,-1])),
    "| min prob:", round(min(submission[,-1]),4),
    "(blend floors it away from 0)\n")

if (!dir.exists("submissions")) dir.create("submissions")
write.csv(submission, "submissions/submission_mnl_covar_blend.csv", row.names=FALSE)
cat("Wrote submissions/submission_mnl_covar_blend.csv (lambda", BEST_LAMBDA, ")\n")

path <- "submissions_log.csv"; log <- read.csv(path, stringsAsFactors=FALSE)
blend_mean <- blend_tbl$mean_ll[blend_tbl$lambda==BEST_LAMBDA]
blend_sd   <- blend_tbl$sd_ll[blend_tbl$lambda==BEST_LAMBDA]
new  <- data.frame(date=as.character(Sys.Date()), model_name="imelda_mnl_covar_blend",
  description=sprintf("Factors MNL + Price x (income,age,miles,night,segment) interactions, predictions blended %.2f toward uniform for 70/30 robustness", BEST_LAMBDA),
  source_citation="own experiment; teammate feedback on covariate interactions; robustness blend from own 100-seed sweep",
  cv_logloss=blend_mean, public_lb_logloss=NA, gap=NA,
  notes=sprintf("resp-split blended mean %.5f sd %.5f (raw covar mean %.5f, factors-only %.5f). Blend lowers sd = safer across the hidden 70/30 slice; also floors probs away from 0.",
                blend_mean, blend_sd, with_int["mean"], no_int["mean"]),
  stringsAsFactors=FALSE)
write.csv(rbind(log, new), path, row.names=FALSE)
cat("Logged imelda_mnl_covar_blend\n")
```

# Notes

- The covariate interactions are the real signal; the blend (lambda=0.08) is a
  robustness wrapper: it barely changes the mean but lowers the sd across splits,
  so public and private (the hidden 70/30) track each other more closely, and it
  floors every probability away from 0 to cap worst-case log-loss spikes.
- `submission_mnl_covar_blend.csv` is the file to upload; it's logged as
  `imelda_mnl_covar_blend`. Fill in public_lb_logloss + gap after Kaggle.
- To extend the model itself: covariate x attribute interactions (e.g. segment x
  safety attrs), or price x categorical covariates as factors — judge every
  addition by resp-split mean +/- sd, keeping only ones that lower mean without
  inflating sd.
- Given flexible models' large public gaps on this data (xgboost ensemble 0.069),
  this interpretable, blended MNL is expected to generalize more reliably.
```
