---
title: "Robustness pass - make one submission slice-insensitive (70/30)"
output: html_document
date: "2026-07-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

You submit ONE prediction file; Kaggle splits it 70/30. You can't touch the
split, but you can make predictions robust so they score similarly on any slice
(public 70% ~ private 30%). This tests three robustness knobs on the covariate
MNL over 100 respondent-level splits, reporting mean AND sd (sd = how much the
score swings across subsets ~ your 70/30 stability proxy):

  1. CLIP  - floor probabilities away from 0 (caps worst-case log-loss spikes)
  2. BLEND - pull predictions gently toward 0.25 uniform (shrinks overconfidence)
  3. TRIM  - keep only the strongest covariate interactions (less overfit)

# Setup

```{r}
library(mlogit)
train <- read.csv("data/train.csv")
test  <- read.csv("data/test.csv")
train$Choice <- max.col(train[, c("Ch1","Ch2","Ch3","Ch4")])
test$Choice  <- 1L

attrs <- c("CC","GN","NS","BU","FA","LD","BZ","FC","FP","RP",
           "PP","KA","SC","TS","NV","MA","LB","AF","HU")
ftr   <- c("NS","BU","FP","SC","MA","LB")

log_loss <- function(actual, pred, eps=1e-15){actual<-as.matrix(actual);pred<-as.matrix(pred)
  pred<-pred/rowSums(pred);pred<-pmin(pmax(pred,eps),1-eps);-mean(rowSums(actual*log(pred)))}
one_hot <- function(v){m<-matrix(0,length(v),4);m[cbind(seq_along(v),v)]<-1;m}

scale_cov <- function(df){df$incomea_s<-df$incomea/1e5;df$agea_s<-df$agea/50
  df$milesa_s<-df$milesa/100;df$nighta_s<-df$nighta/50;df}
# full interaction set, and a TRIMMED set (income + segment only = strongest, most stable)
add_interactions <- function(df, which_set){
  df<-scale_cov(df)
  for(j in 1:4){
    df[[paste0("PxINC",j)]]<-df[[paste0("Price",j)]]*df$incomea_s
    df[[paste0("PxSEG",j)]]<-df[[paste0("Price",j)]]*df$segmentind
    if(which_set=="full"){
      df[[paste0("PxAGE",j)]]<-df[[paste0("Price",j)]]*df$agea_s
      df[[paste0("PxMIL",j)]]<-df[[paste0("Price",j)]]*df$milesa_s
      df[[paste0("PxNGT",j)]]<-df[[paste0("Price",j)]]*df$nighta_s
    }
  }
  df
}
stems_full <- c("PxINC","PxSEG","PxAGE","PxMIL","PxNGT")
stems_trim <- c("PxINC","PxSEG")
```

# Robustness transforms

```{r}
clip_probs  <- function(P, floor=0.02){P<-pmax(P,floor);P/rowSums(P)}
blend_prior <- function(P, lambda=0.05){((1-lambda)*P + lambda*0.25) }  # rows already sum ~1
```

# Fit covariate MNL on a split, return the raw validation prob matrix + truth

```{r}
fit_split <- function(seed, which_set){
  set.seed(seed)
  vc<-sample(unique(train$Case),round(0.20*length(unique(train$Case))))
  fit<-subset(train,!(Case %in% vc)); val<-subset(train,Case %in% vc)
  for(a in ftr) for(j in 1:4){col<-paste0(a,j);tb<-table(fit[[col]])
    keep<-as.numeric(names(tb)[tb>=30]);modal<-as.numeric(names(which.max(tb)))
    fit[[col]][!(fit[[col]]%in%keep)]<-modal; val[[col]][!(val[[col]]%in%keep)]<-modal}
  stems <- if(which_set=="full") stems_full else stems_trim
  fit<-add_interactions(fit,which_set); val<-add_interactions(val,which_set)
  extra<-unlist(lapply(stems,function(s)paste0(s,1:4)))
  vn<-c(names(fit)[4:83],extra)
  Sf<-dfidx(fit,shape="wide",choice="Choice",sep="",varying=which(names(fit)%in%vn),idx=c("No","Case"))
  Sv<-dfidx(val,shape="wide",choice="Choice",sep="",varying=which(names(val)%in%vn),idx=c("No","Case"))
  for(a in ftr){Sf[[a]]<-factor(Sf[[a]]);Sv[[a]]<-factor(Sv[[a]],levels=levels(Sf[[a]]))}
  rhs<-paste(paste(c(attrs,"Price"),collapse="+"),"+",paste(stems,collapse="+"))
  M<-mlogit(as.formula(paste("Choice ~",rhs,"- 1")),data=Sf)
  list(P=predict(M,newdata=Sv), truth=one_hot(val$Choice))
}
```

# Evaluate each robustness setting over 100 seeds

```{r}
seeds  <- 1:100
batches<- split(seeds, ceiling(seq_along(seeds)/20))

# a "setting" = which interaction set + which post-transform
eval_setting <- function(which_set, transform){
  all_ll<-c(); bmeans<-c()
  for(b in batches){
    lls<-sapply(b,function(s){
      out<-tryCatch(fit_split(s,which_set),error=function(e)NULL)
      if(is.null(out)) return(NA)
      P<-transform(out$P)
      log_loss(out$truth,P)
    })
    lls<-lls[!is.na(lls)]; all_ll<-c(all_ll,lls); bmeans<-c(bmeans,mean(lls))
  }
  data.frame(mean_ll=round(mean(all_ll),5), sd=round(sd(all_ll),5),
             se=round(sd(bmeans),5))
}

settings <- list(
  "full_raw"          = list("full", function(P) P),
  "full_clip0.02"     = list("full", function(P) clip_probs(P,0.02)),
  "full_blend0.05"    = list("full", function(P) blend_prior(P,0.05)),
  "full_clip+blend"   = list("full", function(P) clip_probs(blend_prior(P,0.05),0.02)),
  "trim_raw"          = list("trim", function(P) P),
  "trim_clip+blend"   = list("trim", function(P) clip_probs(blend_prior(P,0.05),0.02))
)

res <- do.call(rbind, lapply(names(settings), function(nm){
  s <- settings[[nm]]
  r <- eval_setting(s[[1]], s[[2]]); r$setting <- nm; r
}))
res <- res[order(res$mean_ll), c("setting","mean_ll","sd","se")]
res
```

# Read it

```{r}
cat("mean_ll = expected score; sd = swing across subsets (70/30 stability proxy);\n")
cat("se = precision of the mean over 100 seeds.\n\n")
cat("Pick the setting with the LOWEST mean that also has a SMALL sd.\n")
cat("Clipping/blending usually barely change the mean but SHRINK the sd -> more\n")
cat("stable across whatever 70/30 slice Kaggle draws. That's the robustness win.\n")
write.csv(res, "robustness_pass.csv", row.names=FALSE)
```

# Generate the robustified submission (set winner below)

```{r eval=FALSE}
# choose the winning setting from the table:
WIN_SET   <- "full"                                  # "full" or "trim"
WIN_TRANS <- function(P) clip_probs(blend_prior(P,0.05),0.02)

# refit on all train with level alignment
level_map<-list()
for(a in ftr){allv<-unlist(train[,paste0(a,1:4)]);tb<-table(allv)
  level_map[[a]]<-list(keep=as.numeric(names(tb)[tb>=30]),modal=as.numeric(names(which.max(tb))))}
collapse_rare<-function(df){for(a in ftr)for(j in 1:4){col<-paste0(a,j)
  df[[col]][!(df[[col]]%in%level_map[[a]]$keep)]<-level_map[[a]]$modal};df}

stems <- if(WIN_SET=="full") stems_full else stems_trim
trX<-add_interactions(collapse_rare(train),WIN_SET)
teX<-add_interactions(collapse_rare(test), WIN_SET)
extra<-unlist(lapply(stems,function(s)paste0(s,1:4)))
vn<-c(names(trX)[4:83],extra)
S_all <-dfidx(trX,shape="wide",choice="Choice",sep="",varying=which(names(trX)%in%vn),idx=c("No","Case"))
S_test<-dfidx(teX,shape="wide",choice="Choice",sep="",varying=which(names(teX)%in%vn),idx=c("No","Case"))
for(a in ftr){S_all[[a]]<-factor(S_all[[a]]);S_test[[a]]<-factor(S_test[[a]],levels=levels(S_all[[a]]))}
rhs<-paste(paste(c(attrs,"Price"),collapse="+"),"+",paste(stems,collapse="+"))
M<-mlogit(as.formula(paste("Choice ~",rhs,"- 1")),data=S_all)
P<-WIN_TRANS(predict(M,newdata=S_test))

sub<-data.frame(No=test$No,Ch1=P[,1],Ch2=P[,2],Ch3=P[,3],Ch4=P[,4])
sample_sub<-read.csv("data/sample_submission.csv")
stopifnot(all(names(sub)==names(sample_sub)),all(sub$No==sample_sub$No),!anyNA(sub))
write.csv(sub,"submissions/submission_robust.csv",row.names=FALSE)

log<-read.csv("submissions_log.csv",stringsAsFactors=FALSE)
new<-data.frame(date=as.character(Sys.Date()),model_name="mnl_covar_robust",
  description="Covariate MNL + clip(0.02)+blend(0.05) robustness pass",
  source_citation="own experiment",cv_logloss=NA,public_lb_logloss=NA,gap=NA,
  notes="Robustified for 70/30 stability; see robustness_pass.csv",stringsAsFactors=FALSE)
write.csv(rbind(log,new),"submissions_log.csv",row.names=FALSE)
cat("Wrote submissions/submission_robust.csv\n")
```
```
